Variant calling with transformers

### 2022-05-28 manuscript/manuscript.tex
manuscript

Accurate reconstruction of the DNA sequence in a sample is a fundamental component of modern genomics analysis. Next Generation Sequencing (NGS) workflows typically begin by aligning sequenced fragments to a reference genome, optionally performing some alignment post-processing steps such as identification of duplicated fragments, and then identifying sequence variants relative to the reference genome. The variant identification step is challenging for many reasons, including technical artifacts in the sequencing process, improper alignments due to sequence homology or low sequence complexity, and the stochastic nature of the library preparation and amplification steps [citation here would be nice].

@@

While replacing the statistical and heuristic components of the allele likelihood calculation with a deep learning model has resulted in improved specificity of variant callers, current tools continue to rely on complex, handcrafted statistical machinery to identify possible alleles. Here, we describe a new approach that replaces both the candidate allele generation and the likelihood calculation with a single, learned model. We recast variant detection as a sequence-to-sequence translation problem in which the input sequence is the series of bases aligning to a single genomic location, and the output sequences are the two predicted haplotypes. The sequence-to-sequence translation is performed by a transformer-based deep neural network (Vaswani et al. 2017). We demonstrate that this approach identifies real genomic variants with sensitivity equivalent to current top-performing callers, albeit with slightly lower specificity.

@@

Training tensors were generated by selecting regions 150bp in width from a list of preselected regions, in a manner described below. For each region, all reads overlapping the window were obtained from the corresponding BAM/ CRAM file. If more than a specified maximum number of reads were found (100 for all analyses here), then reads were downsampled to the desired size.  Reads were sorted by the reference coordinate of the first aligned base. For each aligned base in the selected reads nine features were encoded; the first four were the one-hot encoded base call, followed by base quality, two flags indicating if base `consumed' reference base (i.e. was not an insertion) or consumed a sequence read base (was not a deletion), and additional flags indicating sequence read direction and clipping status.  No 'gap' tokens or other special handling was performed for insertions or deletions. In addition to sequenced reads, the first row in each encoded region was the reference sequence. For this special row we inserted a base quality of 100 for every position and did not set any of the other flags. Resulting tensors had dimension $[g, r, 9]$, where $g$ represents genomic position, $r$ represents read, and 9 is the number of features.

@@

To select regions to include for training data, we developed a scheme to sample regions in a biased manner, prioritizing regions containing variants and, especially, regions with multiple or complex variants. Regions of the reference genome overlapping the high-confidence regions from Genome-in-a-Bottle were subdivided into 150bp nonoverlapping windows, and these regions labelled according to the presence of variants. Separate labels were generated for regions conataining a single SNV, deletion, or insertion, as well as regions containing multiple insertion-deletions, or those containing variants intersecting  low-complexity regions. We additionally included `true negative' regions where no known variant was present. The total number of regions generated per sample was 325,000, yielding 5.53M total training regions across our 17 samples.

@@

We use a standard cross-entropy loss function to compare predicted haplotypes to known haplotype sequences for each region, and simply compare each prediction at window index (position) $i$ to the true base at index $i$.  While we obtain satisfactory results with this approach, two issues were apparent. First, predicted haplotype sequences may have inserted or deleted bases relative to the true haplotypes. Because incorrect insertions and deletions cause mismatches at every downstream (higher index) position, such mistakes are penalized more heavily when they occur at lower indexes than higher indexes. Thus the model learns to generate high quality results for the first portion of the window where inaccuracies are punished severely, and is less concerned with accuarcy at high indexes.

A second complexity arises due to ambiguity regarding which haplotype should contain a variant. The two haplotypes present in a sample have no inherent order. In training data variants are assigned arbitrarily to haplotypes, and this randomness in haplotype assignment means model cannot learn to predict which haplotype should contain a given variant. To combat this ambiguity, we calculate loss in both possible configurations - predicted haplotype 0 with training haplotype 0 and 1 with 1, as well as predicted haplotype 0 with training haplotype 1 and 1 with 0. We select the configuration with the lowest loss and backpropogate gradients only for that configuration.

@@

Given a region containing suspected variants, we perform multiple overlapping forward passes of the model with step size $k$. On each forward pass the model produces two predicted haplotypes by taking the base with the high predicted probability at each position. Each haplotype is aligned via Smith-Waterman to the reference sequence, and any mismatching positions are converted to variant calls. For each variant we record the number of windows in which each variant was detected, the number of variants in cis and trans, and the mean probability of the variant bases, and the position of the variant within each window.

@@

Our analysis demonstrates that the model learns many features of modern, state of the art variant callers. For instance, our model learns to reconstruct large, complex variants in the presence of ambiguous mapping and extensive soft clipping. In addition, the model learns to employ local context and read-specific features to predict phase

 Loss functions, there's probably something better. The Smith Waterman loss sounds really cool, but didn't lead to high sensitivity in our hands.

@@

>>>>

Accurate reconstruction of the DNA sequence in a sample is a fundamental component of modern genomics analysis. Next Generation Sequencing (NGS) workflows typically begin by aligning sequenced fragments to a reference genome, optionally performing some alignment post-processing steps such as identification of duplicated fragments, and then identifying sequence variants relative to the reference genome. The variant identification step is challenging for many reasons, including technical artifacts in the sequencing process, improper alignments due to sequence homology or low sequence complexity, and the stochastic nature of the library preparation and amplification steps (Li 2014).

@@

While replacing the statistical and heuristic components of the allele likelihood calculation with a deep learning model has resulted in improved specificity of variant callers, current tools continue to rely on complex, handcrafted statistical machinery to identify possible alleles. Here, we describe a new approach that replaces both the candidate allele generation and the likelihood calculation with a single, learned model. We recast variant detection as a sequence-to-sequence translation problem in which the input sequence is the series of bases aligning to a single genomic location, and the output sequences are the two predicted haplotypes. The sequence-to-sequence translation is performed by a transformer-based deep neural network (Vaswani et al. 2017). We demonstrate that this approach identifies real genomic variants with sensitivity similar to current top-performing callers, albeit with lower specificity.

@@

Training tensors were generated by selecting regions 150bp in width from a list of preselected regions, in a manner described below. For each region, all reads overlapping the window were obtained from the corresponding BAM/ CRAM file. If more than a specified maximum number of reads were found (100 for all analyses here), then reads were downsampled to the desired size.  Reads were sorted by the reference coordinate of the first aligned base. For each aligned base in the selected reads nine features were encoded; the first four were the one-hot encoded base call, followed by base quality, two flags indicating if base `consumed' reference base (i.e. was not an insertion) or consumed a sequence read base (was not a deletion), and additional flags indicating sequence read direction and clipping status.  No 'gap' tokens or other special handling was performed for insertions or deletions. In addition to sequenced reads, the first row in each encoded region was the reference sequence. For this special row we inserted a base quality of 100 for every position and did not set any of the other flags. Resulting tensors had dimension $[150, 100, 9]$, where the indices are for the genomic position, read, and feature respectively.

@@

To select regions to include for training data, we developed a scheme to sample regions in a biased manner, prioritizing regions containing variants and, especially, regions with multiple or complex variants. Regions of the reference genome overlapping the high-confidence regions from Genome-in-a-Bottle were subdivided into 150bp nonoverlapping windows, and these regions labelled according to the presence of variants. Separate labels were generated for regions conataining a single SNV, deletion, or insertion, as well as regions containing multiple insertion-deletions, or those containing variants intersecting  low complexity regions. Low complexity regions were obtained from Genome-in-a-Bottle stratification BED files. We additionally included `true negative' regions where no known variant was present. The total number of regions generated per sample was 325,000, yielding 5.53M total training regions across our 17 samples.

@@

We use a standard cross-entropy loss function to compare predicted haplotypes to known haplotype sequences for each region, and simply compare each prediction at window index (position) $i$ to the true base at index $i$. A complexity arises due to ambiguity regarding which haplotype should contain a variant. The two haplotypes present in a sample have no inherent order. In training data variants are assigned arbitrarily to haplotypes, and this randomness in haplotype assignment means the model cannot learn to predict which haplotype should contain a given variant. To combat this ambiguity, we calculate loss in both possible configurations - predicted haplotype 0 with training haplotype 0 and 1 with 1, as well as predicted haplotype 0 with training haplotype 1 and 1 with 0. We select the configuration with the lowest loss and backpropogate gradients only for that configuration.

@@

Given a region containing suspected variants, we perform multiple overlapping forward passes of the model with step size $k$, where $k=50$ for the results reported here. On each forward pass the model produces two predicted haplotypes by taking the base with the high predicted probability at each position. Each haplotype is aligned via Smith-Waterman to the reference sequence, and any mismatching positions are converted to variant calls. For each variant we record the number of windows in which each variant was detected, the number of variants in cis and trans, and the mean probability of the variant bases, and the position of the variant within each window.

@@

Our analysis demonstrates that the model learns many features of modern, state of the art variant callers. For instance, our model learns to reconstruct large, complex variants in the presence of ambiguous mapping and extensive soft clipping. In addition, the model learns to employ local context and read-specific features to predict phase and genotype accurately.

In this work we have used a standard cross-entropy loss function that simply compares each prediction at base $i$ to the label at base $i$. While training with this loss yields accurate results, it has two significant shortcomings. First, the impact of an insertion / deletion (indel) error depends on its position in the region, with indels at lower indices leading to a greater number of mismatches and high loss compared to indels at higher indices. Secondly, indel errors ar punished more severely than similar SNV errors. We hypothesize the second fact explains the lower performance of our model on SNV calls compared to indels. Alternative, alignment-based loss functions such as those described by Baid et al. (2021) and Petti et al. (2021) may provide a path forward, however the Petti implementation lead to lower sensitivity in our experiments.

@@

\vspace{8pt}
Li, Heng. "Toward better understanding of artifacts in variant calling from high-coverage samples." Bioinformatics 30.20 (2014): 2843-2851.
